---
title: "431 Class 03"
author: "Thomas E. Love, Ph.D."
date: "2022-09-06"
format:
  revealjs: 
    theme: simple
    self-contained: true
    slide-number: true
    preview-links: auto
    logo: 431-class-foot2.png
    footer: "431 Class 03 | 2022-09-06 | https://thomaselove.github.io/431-2022/"
---

## Today's Agenda {.smaller}

- Work in R with a familiar data set (the 15 question survey from Class 02)
- Open RStudio, load in some data and a template to write R Markdown code
  - We'll do a little typing into the template today, but just a little.
    - We'll then look at the completed R Markdown document.
    - We'll also inspect and knit the R Markdown file after all of the code is included.
  - Then we'll start over again with the slides.
- These slides walk through everything in that R Markdown document

::: aside
Version `r Sys.time()`
:::

## Today's Files

From our [431-data page](https://github.com/THOMASELOVE/431-data), or our [Class 03 README (data folder)](https://github.com/THOMASELOVE/431-classes-2022/tree/main/class03/data), you should find:

- `431-first-r-template.Rmd`
- `quick_survey_2022.csv`

and

- `431-class03-all-code.Rmd`

in addition to the usual slide materials.

## Today's Plan

We're using R Markdown to gather together into a single document:

- the code we build, 
- text commenting on and reacting to that code, and 
- the output of the analyses we build.

Everything in these slides is also going into our R Markdown file.


## Load packages and set theme

```{r}
#| echo: true
#| message: false

library(janitor)
library(patchwork)
library(tidyverse)

theme_set(theme_bw())
```

Loading packages in R is like opening up apps on your phone. We need to tell R that, in addition to the base functions available in the software, we also have other functions we want to use. 

- Why are we loading these packages, in particular?

## On the tidyverse meta-package {.smaller}

- The most important package is actually a series of packages called the `tidyverse`, which we'll use in every R Markdown file we create this semester. 
    - The `tidyverse` includes several packages, all developed (in part) by Hadley Wickham, Chief Scientist at RStudio.
    - `dplyr` is our main package for data wrangling, cleaning and transformation
    - `ggplot2` is our main visualization package we'll use for visualization
    - other `tidyverse` packages help import data, work with factors and other common activities.
    
## More on today's packages

- The `janitor` package has some tools for examining, cleaning and tabulating data (including `tabyl()` and `clean_names()`) that we'll use regularly.
- The `patchwork` package will help us show multiple `ggplots` together.
- It's helpful to load the `tidyverse` package last.

## Today's Data

Our data come from the Quick 15-item Survey we did in Class 02 ([pdf in Class 02 README](https://github.com/THOMASELOVE/431-classes-2022/tree/main/class02)), which we've done (in various forms) since 2014. 

- A copy of these data (in .csv format) is on our [431-data page](https://github.com/THOMASELOVE/431-data), and also linked on our [Class 03 README](https://github.com/THOMASELOVE/431-classes-2022/tree/main/class03).

We'll tackle several exploratory questions of interest...

## Our Questions of Interest

1. What is the distribution of pulse rates among students in 431 since 2014?
2. Does the distribution of student heights change materially over time?
3. Is a Normal distribution a good model for our data?
4. Do taller people appear to have paid less for their most recent haircut?
5. Do students have a more substantial tobacco history if they prefer to speak English or a language other than English?

## Read in data from `.csv` file

```{r}
#| echo: true
quicksur_raw <- 
  read_csv("c03/data/quick_survey_2022.csv", show_col_types = FALSE) |>
  clean_names()
```

- Note the `<-` assignment arrow to create `quicksur_raw`
- Here, we use `read_csv` to read in data from the `c03/data` subfolder of my R project directory which contains the `quick_survey_2022.csv` file from our [431-data page](https://github.com/THOMASELOVE/431-data).
- We use `show_col_types = FALSE` to suppress some unnecessary output describing the column types
- We use `clean_names()` from the janitor package
- Note the use of the pipe `|>` to direct the information flow

## What is the result?

```{r}
#| echo: true
quicksur_raw
```

## A more detailed look?

```{r}
#| echo: true
glimpse(quicksur_raw)
```

## Counting Categories

```{r}
#| echo: true
quicksur_raw |> count(glasses)
```

```{r}
#| echo: true
quicksur_raw |> count(glasses, english)
```

## Favorite Color in 2022?

```{r}
#| echo: true
quicksur_raw |>
    filter(year == "2022") |>
    tabyl(favcolor) |>
    adorn_pct_formatting()
```

## Using `summary()` on Quantities

```{r}
#| echo: true
quicksur_raw |> 
  select(love_htcm, haircut, height_in, lastsleep) |>
  summary()
```

- Numerical summaries (five quantiles, plus the mean) for:
  - your guess of my height (in cm), the price of your last haircut, your height (in inches), and the hours of sleep you got last night
- How many observations are available for these measures?

# Manage the data into `qsdat`

## Recall our Questions of Interest

1. What is the distribution of pulse rates among students in 431 since 2014?
2. Does the distribution of student heights change materially over time?
3. Is the Normal distribution a good model for our data?
3. Do taller people appear to have paid less for their most recent haircut?
4. Do students have a more substantial tobacco history if they prefer to speak English or a language other than English?

## Variables we'll look at closely today {.smaller}

To address our Questions of Interest, we need these seven variables in our analytic data frame (tibble.)

-   `student`: student identification (numerical code)
-   `year`: indicates year when survey was taken (August)
-   `english`: y = prefers to speak English, else n
-   `smoke`: 1 = never smoker, 2 = quit, 3 = current
-   `pulse`: pulse rate (beats per minute)
-   `height_in`: student's height (in inches)
-   `haircut`: price of student's last haircut (in \$)

## Select our variables

```{r}
#| echo: true
qsdat <- quicksur_raw |>
    select(student, year, english, smoke, 
           pulse, height_in, haircut)
```

- The `select()` function chooses the variables (columns) we want to keep in our new tibble called `qsdat`.
- What should the result of this code look like?

## What do we have now?

```{r}
#| echo: true
qsdat
```

## Initial Numeric Summaries

- Is everything the "type" of variable it should be? 
- Are we getting the summaries we want?

```{r}
#| echo: true
summary(qsdat)
```

## What should we be seeing?

- Categorical variables should list the categories, with associated counts. 
  - To accomplish this, the variable needs to be represented in R with a `factor`, rather than as a `character` or `numeric` variable.
- Quantitative variables should show the minimum, median, mean, maximum, etc.

```{r}
#| echo: true
names(qsdat)
```

## Change categorical variables to factors

We want the `year` and `smoke` information treated as categorical, rather than as quantitative, and the `english` information as a factor, too. Also, do we want to summarize the student ID codes?

- We use the `mutate()` function to help with this.

```{r}
#| echo: true
qsdat <- qsdat |>
    mutate(year = as_factor(year),
           smoke = as_factor(smoke),
           english = as_factor(english),
           student = as.character(student))
```

- Note that it's `as_factor()` but `as.character()`. Sigh.

## Next step: Recheck the summaries and do range checks

-   Do these summaries make sense?
-   Are the minimum and maximum values appropriate?
-   How much missingness are we to deal with?

## Now, how's our summary?

```{r}
#| echo: true

summary(qsdat)
```

- Some things to look for appear on the next slide.

## What to look for...

- Are we getting counts for all variables that are categorical?
    - Do the category levels make sense?
- Are we getting means and medians for all variables that are quantities?
    - Do the minimum and maximum values make sense for each of these quantities?
- Which variables have missing data, as indicated by `NA's`?

## The summary for `year` is an issue

- Just to fill in the gap left by the `summary()` result, how many students responded each year?

```{r}
#| echo: true
qsdat |> tabyl(year) |> adorn_totals() |> adorn_pct_formatting()
```

## This is how far we got in Class 03. 

See the Class 04 slides for the remainder of the materials originally posted here.

## Session Information

Don't forget to close your file with the session information.

```{r}
#| echo: true
sessionInfo()
```